---
title: "Using Coala for Approximate Bayesian Computation"
author: "Paul Staab"
date: "coala `r packageVersion('coala')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using coala for ABC}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

[Approximate Bayesian Computation (ABC)](http://en.wikipedia.org/wiki/Approximate_Bayesian_computation)
is a simulation based method for Bayesian inference. It is commonly used in
evolutionary biology to estimate parameters of e.g. demographic models.
Coala makes it easy to conduct the simulations for an ABC analysis and works
well together with the `abc` package for doing the estimation. 
To demonstrate the principle, we will estimate the parameter of an 
over-simplified toy model. 

Lets assume we have 50 genetic loci from 10 individuals from a panmictic 
population. We'll use the site frequency spectrum of the data as a summary
statistic to estimate the scaled mutation rate `theta`. Lets assume the
frequency spectrum we get from the data is

```{r sfs}
sfs <- c(112, 57, 24, 34, 16, 29, 8, 10, 15)
```



We can now use coala to setup the model:

```{r model setup}
library(coala)
model <- coal_model(10, 50) +
  feat_mutation(par_prior("theta", runif(1, 1, 5))) +
  sumstat_sfs()
```

Note that we used `par_prior` to set a flat uniform prior between 1 and 5 for
`theta`. 

We can now do easily simulate the model:

```{r simulate, cache=TRUE}
sim_data <- simulate(model, nsim = 2000, seed = 17)
```

For this toy model, we did just 2000 simulations (to keep the time for
building this document within a reasonable range). A real analysis will need
many more!


We now need to prepare the simulation data for the `abc` package:

```{r}
# Getting the parameters
sim_pars <- matrix(sapply(sim_data, function(x) x$pars), 2000, 1)
colnames(sim_pars) <- "theta"
head(sim_pars, n = 3)

# Getting the summary statistics
sim_sumstats <- t(sapply(sim_data, function(x) x$sfs))
colnames(sim_sumstats) <- paste0("S", 1:9)
head(sim_sumstats, n = 3)
```


And can now estimate `theta`:

```{r abc, fig.align="center", fig.width=5}
suppressPackageStartupMessages(library(abc))
posterior <- abc(sfs, sim_pars, sim_sumstats, 0.05, method = "rejection")
hist(posterior, breaks = 10)
```

Due to the low number of simulations we get a relatively flat posterior,
but we still get an idea what the value for `theta` might be.

