---
title: "Importing Genetic Data into Coala"
author: "Paul Staab"
date: "coala `r packageVersion('coala')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

While `coala` primiary focuses on simulation of data, it also offers
to calculate summary statsitcs from real data. This is particular useful
when comparing the statistics of real and simulated data.

Rather than offering function to import data directly, coala can convert
the internal formats of other packages into its own format. Currently, only
`PopGenome` is supported, but we plan to also support `ape` and `pegas` in the
future.


Importing Data from PopGenome
-----------------------------

PopGenome offers functions for reading different data formats, including `vcf` 
and `fasta`. Please refer to its documentation for instructions. As example,
we will read in one short fasta file that is included in coala:

```{r import_data, results="hide"}
suppressPackageStartupMessages(library(PopGenome))
fasta <- system.file("example_fasta_files", package = "coala")
data_pg <- readData(fasta, progress_bar_switch = FALSE)
data_pg <- set.outgroup(data_pg, c("Individual_Out-1", "Individual_Out-2"))
individuals <- list(paste0("Individual_1-", 1:5), paste0("Individual_2-", 1:5))
data_pg <- set.populations(data_pg, individuals)
```

We can now convert `data_pg` using the `as.segsites` function: 

```{r}
library(coala)
segsites <- as.segsites(data_pg)
```

Now, we can calculate summary statistics using `calc_sumstats_from_data`:
```{r calc_sumstats_1}
model <- coal_model(c(5, 5, 2), 1, 25) + 
  feat_mutation(5) +
  feat_outgroup(3) +
  sumstat_sfs(population = 1)
stats <- calc_sumstats_from_data(model, segsites)
stats$sfs
```


Alternatively, it is also possible to pass the `data_pg` object directly to
`calc_sumstats_from_data`:

```{r calc_sumstats_2}
stats <- calc_sumstats_from_data(model, data_pg)
stats$sfs
```

Please refer to the help pages for `as.segsites` and `calc_sumstats_from_data`
for additional information.
