---
title: "Introduction to coalsimr"
author: "Paul Staab"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Coalescent simulation referrers to the idea to simulate the evolution of 
biological sequences like DNA by tracing their ancestry back in time. The
`coalsimr` package is an interface for calling a number of commonly used
coalescent simulators from R.

# Creating a model
A basic model can be created using the function `coal_model`:

```{r}
library(coalsimr)
model <- coal_model(sample_size = 5, loci_number = 1)
```

This creates a basic model of a population of constant size.
Simulated is one genetic locus in 5 individuals taken from the population. 

Printing the model gives a short summary of this content:
```{r}
print(model)
```

simulate trees.


# Adding features

mutations, simulate dna

two populations, migration

# Parameters

# Loci

# Summary statistics
