---
title: "Introduction to coalsimr"
author: "Paul Staab"
date: "coalsimr `r packageVersion('coalsimr')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Coalescent simulation referrers to the idea to simulate the evolution of 
biological sequences like DNA by tracing their ancestry back in time. The
`coalsimr` package is an interface for calling a number of commonly used
coalescent simulators from R.

# Creating a model
A basic model can be created using the function `coal_model`:

```{r}
library(coalsimr)
model <- coal_model(sample_size = 5, loci_number = 1)
```

This creates a basic model of a population of constant size.
Simulated is one genetic locus in 5 individuals taken from the population. 

Printing the model gives a short summary of this content:
```{r}
print(model)
```

simulate trees. Models consist of 

* _features_, which represent evolutionary forces and events present in the 
  model, 
* _parameters_, which are the model parameters, 
* _loci_, which describe the genetic regions that are simualted and 
* _summary statistics_, which describe the format of the simulated data.

In our basic model, the sampling of 5 individuals the only feature. 

simulated trees

# Adding features

mutations, simulate dna

two populations, migration

# Parameters

# Loci

# Summary statistics
